---
title: "Assignment 10: Data Scraping"
author: "Sky Volz"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)
library(here)
library(cowplot)
here()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Set the year to 2024
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
the_website <- 
  read_html("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024")

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * Contact Person
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "Reginald Hicks", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
the_system_name <- the_website %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()

the_contact_person <- the_website %>% 
  html_nodes("div+ table tr:nth-child(4) td:nth-child(2)") %>% html_text()

the_ownership <- the_website %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()

maximum_day_use <- the_website %>% 
  html_nodes("th~ td+ td") %>% html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
max_day_df <- data.frame(
  "contact.person" = the_contact_person,
  "ownership" = the_ownership,
  "system.name" = the_system_name,
  "max.day.use.MGD" = as.numeric(maximum_day_use),
  "month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
  "year" = 2024
)

max_day_df <- max_day_df %>% 
  mutate(date = my(paste(month,"-",year)))

#5 
max_day_plot <- ggplot(max_day_df, aes(x = date,y=max.day.use.MGD)) +
  geom_line() + 
  labs(x="Month", y="Maximum Daily Use (MGD)") + scale_x_date(
    breaks = "1 month", date_labels = "%b")

print(max_day_plot)

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs, "PWSID" and "year", that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
scrape.it <- function(PWSID, year){
  base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid='
  scrape_url <- paste0(base_url, PWSID, "&year=", year)
  the_website <- read_html(scrape_url)

  #setting element address variables
  the_contact_person <- 'div+ table tr:nth-child(4) td:nth-child(2)'
  the_ownership <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  the_system_name <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  maximum_day_use <- 'th~ td+ td'
  #scraping the data items
  the_contact_person <- the_website %>% html_nodes(the_contact_person) %>% html_text()
  the_ownership <- the_website %>%   html_nodes(the_ownership) %>%  html_text()
  the_system_name <- the_website %>% html_nodes(the_system_name) %>% html_text()
  maximum_day_use <- the_website %>% html_nodes(maximum_day_use) %>% html_text()
  #creating data frame
   df_day_use <- data.frame(
    contact.person = the_contact_person,
    ownership = the_ownership,
    system.name = the_system_name,
    max.day.use.MGD = maximum_day_use,
    month = c(1,5,9,2,6,10,3,7,11,4,8,12),
    PWSID = PWSID,
    year = year
  )

  return(df_day_use)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
df <- scrape.it("03-32-010",2020)

#making sure the max use value and the month are read as numeric and not characters
df <- df %>%
  mutate(
    max.day.use.MGD = as.numeric(gsub(",", "", max.day.use.MGD)),
    month = as.numeric(month)
  )

df <-  df %>% 
  mutate(date = my(paste(month,"-",year)))

day_use_plot <- ggplot(df, aes(x = date,y=max.day.use.MGD, group = 1)) +
  geom_line() + 
  labs(x="Month", y="Maximum Daily Use (MGD)", title="Durham, 2020") +
  scale_x_date(
    breaks = "1 month", date_labels = "%b")

print(day_use_plot)
str(df)
```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
df2 <- scrape.it("03-92-020",2020)

#making sure the max use value and the month are read as numeric and not characters
df2 <- df2 %>%
  mutate(
    max.day.use.MGD = as.numeric(gsub(",", "", max.day.use.MGD)),
    month = as.numeric(month)
  )

df2 <-  df2 %>% 
  mutate(date = my(paste(month,"-",year)))

day_use_plot2 <- ggplot(df2, aes(x = month,y=max.day.use.MGD, group = 1)) +
  geom_line() + 
  labs(x="Month", y="Maximum Daily Use (MGD)", title="Cary, 2020") +
  scale_x_date(
    breaks = "1 month", date_labels = "%b")

str(df2)

compare <- ggplot() +
  geom_line(
    data=df, aes(x = date,y=max.day.use.MGD, group=1, color = "Durham")) +
  geom_line(
    data=df2, aes(x = date,y=max.day.use.MGD, group=1, color="Cary")) +
  labs(
    x="Month", y="Maximum Daily Use (MGD)", 
    title = "Water Withdrawl for Cary and Durham, NC") +
  scale_x_date(breaks = "1 month", date_labels = "%b") +
  theme(legend.position = "right")

print(compare)
```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "06_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs (where one of the two inputs does not change). Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
the_years <- 2018:2023
cary_id <- '03-92-020'
cary_withdrawals <- map2(rep(cary_id, length(the_years)),the_years, scrape.it) %>%
  bind_rows()

#making sure the max daily use values are read as numeric and not characters
cary_withdrawals <- cary_withdrawals %>%
  mutate(max.day.use.MGD = as.numeric(max.day.use.MGD), month=as.numeric(month))

cary_withdrawals <- cary_withdrawals %>% 
  mutate(date = my(paste(month,"-",year)))

cary_plot <- ggplot(cary_withdrawals, aes(x = date,y=max.day.use.MGD)) +
  geom_line() + 
  geom_smooth(method="loess", se=FALSE) +
  labs(x="Year", y="Maximum Daily Use (MGD)", 
       title = "Cary's Maximum Daily Withdrawal for 2018-2024") 

print(cary_plot)
```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer:There looks to be a slight increase over time in water used, as the smoothed line is increasing somewhat as time goes one. However, this increase is very slight.
>
